Plug-and-play dual-tree algorithm runtime analysis

### 2015-01-08 bounds-files/sec03-approx-kde.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

In the case of an infinite-tailed kernel $\mathcal{K}(\cdot, \cdot)$, the
computation cannot be accelerated; thus, attention has turned towards tractable
approximation schemes.  Two simple schemes for the approximation of $f^*(p_q)$
are well-known: {\it absolute value approximation} and {\it relative value
approximation}.  Absolute value approximation requires that each density
estimate $f(p_q)$ is within $\epsilon$ of the true estimate $f^*(p_q)$:

\begin{equation}
\label{eqn:ava}
| f(p_q) - f^*(p_q) | < N \epsilon.
\end{equation}

@@

\begin{equation}
\label{eqn:rva}
\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
\end{equation}

@@

A tree-independent algorithm for solving approximate kernel density estimation
with absolute value approximation under the previous assumptions on th kernel
is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
correctness proof can be found in][]{curtin2013tree}.  The list $f_p$ holds
partial kernel density estimates for each query point, and the list $f_n$ holds
partial kernel density estimates for each query node.  At the beginning of the
dual-tree traversal, the lists $f_p$ and $f_n$, which are both of size $O(N)$,
are each initialized to 0.  As the traversal proceeds, node combinations are
pruned if the difference between the maximum kernel value
$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r))$ and the minimum kernel
value $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$ is sufficiently
small (line \ref{alg:ava-kde-prune}).  If the node combination can be pruned,
then the partial node estimate is updated (line \ref{alg:ava-kde-update}).  When
node combinations cannot be pruned, \texttt{BaseCase()} may be called, which
simply updates the partial point estimate with the exact kernel evaluation (line
\ref{alg:kde-bc-update}).

After the dual-tree traversal, the actual kernel density estimates $f$
must be extracted.  This can be done by traversing the query tree and
calculating $f(p_q) = f_p(p_q) + \sum_{\mathscr{N}_i \in T}
f_n(\mathscr{N}_i)$, where $T$ is the set of nodes in $\mathscr{T}_q$ that
have $p_q$ as a descendant.  %It is possible to implement this in a way where
Each query node needs to be visited only once.

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some $\epsilon$ error parameter is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
defined as in Lemma \ref{lem:extcase3}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, we will bound $|R^*|$ using techniques related to those used by
\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.  Second,
bound $R^*$ when $s_r^{\max}$ is larger.

The \texttt{Score()} function is such that any node in $|R^*|$ for a given query
node $\mathscr{N}_q$ obeys

@@

Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to our assumptions, the
maximum value of $\mathcal{K}'(\cdot)$ is
$\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  Further,
$\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le (-\epsilon /
\mathcal{K}'(h))$.  Note that

@@

Therefore, $R^* = \emptyset$ when
%\begin{eqnarray}
$2^{s_r^{\max} + 2} \le -\epsilon / \mathcal{K}'(h)$, or when
%\begin{equation}
$s_r^{\max} \le \log_2( -\epsilon / \mathcal{K}'(h) ) - 2$.
%\end{equation}
%
Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
\mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
alternately stated, this means that
%\begin{equation}
$d(p_q, p_r) < \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}$.
%\end{equation}
Therefore, bounding the number of points in the set
%\begin{equation}
$B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$
%\end{equation}
is sufficient to bound $|R^*|$.  For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \lceil \log_2
\omega \rceil}$.

@@

For the Gaussian kernel with bandwidth $\sigma$ defined by $\mathcal{K}_g(d) =
\exp(-d^2 / (2 \sigma^2))$, $\zeta$ does not
depend on the kernel bandwidth; only
the approximation parameter $\epsilon$.  For this kernel, $h = \sigma$ and
therefore $-\mathcal{K}'_g(h) = \sigma^{-1} e^{-1 / 2}$.  Additionally,
$\mathcal{K}_g^{-1}(\epsilon) = \sigma \sqrt{2 \ln (1 / \epsilon)}$.  This means
that for the Gaussian kernel, $\zeta = \sqrt{(-2 \ln \epsilon) / (e
\epsilon^2)}$.  Again, as $\epsilon \to 0$, the
runtime diverges; however, note that there is no dependence on the kernel
bandwidth $\sigma$.  For a reasonably chosen $\epsilon = 0.05$, the runtime is
approximately $O(c_r^{8.89} (N + \theta))$; for $\epsilon = 0.01$, the runtime
is approximately $O(c_r^{11.52} (N + \theta))$.  For very small $\epsilon =
0.00001$, the runtime is approximately $O(c_r^{22.15} (N + \theta))$.

To understand the lack of dependence on kernel bandwidth more intuitively,
consider that as the kernel bandwidth increases, two things happen: {\it (a)}
the reference set $R$ becomes empty at larger scales, and {\it (b)}
$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
These effects are opposite, and for the Gaussian kernel they cancel each other
out, giving the same bound regardless of bandwidth.

@@

@@

\begin{equation}
f^*(p_q) \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
\end{equation}

This is clearly true, because each point in $S_r$ must contribute more than
$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
the relative approximation condition in Equation \ref{eqn:rva}:

@@

\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
\in S_r} d(p_q, p_r))$.  This allows us to create a \texttt{Score()} algorithm,
given in Algorithm \ref{alg:kde_rva_score}.

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ both of size $O(N)$, it is possible to perform relative value
approximate kernel density estimation (satisfying the condition of Equation
\ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant $c_r$ of
$S_r$ is not dependent on $N$.
\end{thm}

>>>>

In the case of an infinite-tailed kernel $\mathcal{K}(\cdot, \cdot)$, the
exact computation cannot be accelerated; thus, attention has turned towards
tractable approximation schemes.  Two simple schemes for the approximation of
$f^*(p_q)$ are well-known: {\it absolute value approximation} and {\it relative
value approximation}.  Absolute value approximation requires that each density
estimate $f(p_q)$ is within $\epsilon$ of the true estimate $f^*(p_q)$:

\begin{equation}
\label{eqn:ava}
| f(p_q) - f^*(p_q) | < \epsilon \; \; \forall p_q \in S_q.
\end{equation}

@@

\begin{equation}
\label{eqn:rva}
\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < \epsilon \; \; \forall p_q \in S_q.
\end{equation}

@@

A tree-independent algorithm for solving approximate kernel density estimation
with absolute value approximation under the previous assumptions on the kernel
is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
correctness proof can be found in][]{curtin2013tree}.  The list $f_p$ holds
partial kernel density estimates for each query point, and the list $f_n$ holds
partial kernel density estimates for each query node.  At the beginning of the
dual-tree traversal, the lists $f_p$ and $f_n$, which are both of size $O(N)$,
are each initialized to 0.  As the traversal proceeds, node combinations are
pruned if the difference between the maximum kernel value
$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r))$ and the minimum kernel
value $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$ is sufficiently
small (line \ref{alg:ava-kde-prune}).  If the node combination can be pruned,
then the partial node estimate is updated (line \ref{alg:ava-kde-update}).  When
node combinations cannot be pruned, \texttt{BaseCase()} may be called, which
simply updates the partial point estimate with the exact kernel evaluation (line
\ref{alg:kde-bc-update}).

After the dual-tree traversal, the actual kernel density estimates $f$
must be extracted.  This can be done by traversing the query tree and
calculating $f(p_q) = f_p(p_q) + \sum_{\mathscr{N}_i \in T}
f_n(\mathscr{N}_i)$, where $T$ is the set of nodes in $\mathscr{T}_q$ that
have $p_q$ as a descendant.  %It is possible to implement this in a way where
Each query node needs to be visited only once to perform this calculation; it
may therefore be accomplished in $O(N)$ time.

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous subsection.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some error parameter $\epsilon$ is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
%\end{equation}
with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
defined as in Lemma \ref{lem:extcase3}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, we will bound $|R^*|$ using techniques related to those used by
\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
we show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.
Second, we bound $R^*$ when $s_r^{\max}$ is larger.

The \texttt{Score()} function is such that any node in $R^*$ for a given query
node $\mathscr{N}_q$ obeys

@@

Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to our assumptions, the
maximum value of $\mathcal{K}'(\cdot)$ is
$\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  This allows
us to say that $\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le
(-\epsilon / \mathcal{K}'(h))$.  Note that

@@

Therefore, $R^* = \emptyset$ when
%\begin{eqnarray}
$2^{s_r^{\max} + 2} \le -\epsilon / \mathcal{K}'(h)$, or when
%\begin{equation}
$s_r^{\max} \le \log_2( -\epsilon / \mathcal{K}'(h) ) - 2$.
%\end{equation}
%
Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
\mathcal{K}'(h) ) - 2$.  Because of the pruning rule, for any $\mathscr{N}_r \in
R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
we may refactor this by applying definitions to show
%\begin{equation}
$d(p_q, p_r) < \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}$.
%\end{equation}
Therefore, bounding the number of points in the set
%\begin{equation}
$B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$
%\end{equation}
is sufficient to bound $|R^*|$.  For notational convenience, define $\omega =
(\mathcal{K}^{-1}(\epsilon) / 2^{s_r^{\max} + 1}) + 1$, and the statement may be
more concisely written as $B_{S_r}(p_q, \omega 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$.

Using Lemma \ref{lem:packing} with $\delta = 2^{s_r^{\max}}$ and $\rho = 2
\omega$ gives $|R^*| = c_r^{3 + \lceil \log_2 \omega \rceil}$.

@@

For the Gaussian kernel with bandwidth $\sigma$ defined by $\mathcal{K}_g(d) =
\exp(-d^2 / (2 \sigma^2))$, $\zeta$ does not
depend on the kernel bandwidth; only
the approximation parameter $\epsilon$.  For this kernel, $h = \sigma$ and
therefore $-\mathcal{K}'_g(h) = \sigma^{-1} e^{-1 / 2}$.  Additionally,
$\mathcal{K}_g^{-1}(\epsilon) = \sigma \sqrt{2 \ln (1 / \epsilon)}$.  This means
that for the Gaussian kernel, $\zeta = \sqrt{(-2 \ln \epsilon) / (e
\epsilon^2)}$.  Again, as $\epsilon \to 0$, the
runtime diverges; however, note that there is no dependence on the kernel
bandwidth $\sigma$.  To demonstrate the relationship of runtime to $\epsilon$,
see that for a reasonably chosen $\epsilon = 0.05$, the runtime is approximately
$O(c_r^{8.89} (N + \theta))$; for $\epsilon = 0.01$, the runtime is
approximately $O(c_r^{11.52} (N + \theta))$.  For very small $\epsilon =
0.00001$, the runtime is approximately $O(c_r^{22.15} (N + \theta))$.

@@

To understand the lack of dependence on kernel bandwidth more intuitively,
consider that as the kernel bandwidth increases, two things happen: {\it (a)}
the reference set $R$ becomes empty at larger scales, and {\it (b)}
$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
These effects are opposite, and for the Gaussian and Laplacian kernels they
cancel each other out, giving the same bound regardless of bandwidth.

@@

\begin{equation}
|f^*(p_q)| \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
\end{equation}

This is clearly true: each point in $S_r$ must contribute more than
$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
the relative approximation condition in Equation \ref{eqn:rva}:

@@

\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
\in S_r} d(p_q, p_r))$.  Assuming we have some estimate $\mathcal{K}^{\max}$,
this allows us to create a \texttt{Score()} algorithm, given in Algorithm
\ref{alg:kde_rva_score}.

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the same
assumptions as Theorem \ref{thm:kde-bound}.  Then, given a query set $S_q$ and a
reference set $S_r$ both of size $O(N)$, it is possible to perform relative
value approximate kernel density estimation (satisfying the condition of
Equation \ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant
$c_r$ of $S_r$ is not dependent on $N$.
\end{thm}

### 2015-01-08 bounds-files/sec04-range-search.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

In the range search problem, the task is to find the set of reference points
%\begin{equation}
$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
%\end{equation}
for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|S[p_q]|$, is desired.
Our proof works for both of these algorithms
identically, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

@@

Small $| S_{\max} |$ implies an easy problem; large $| S_{\max} |$ implies a
difficult problem.  For bounding the running time of range search, we require
one more definition.

@@

Using these notions, we may now bound the running time of range search.

@@

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
proofs, then, our only task is to bound the maximum size of the reference set,
$|R^*|$.

@@

This set is necessarily a subset of $S^{\alpha}[p_q]$; by
definition, the number of points in this set is bounded above by $|S_{\max}| +
C$.  We may then conclude that $|R^*| \le |S_{\max}| + C$.  By taking the
maximum of the sizes of $|R^*|$ in both cases above, we obtain the statement of
the theorem.
\end{proof}

>>>>

In the range search problem, the task is to find the set of reference points

\begin{equation}
$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
\end{equation}

for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|S[p_q]|$, is desired.
Our proof works for both of these algorithms
similarly, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

@@

Small $| S_{\max} |$ implies an easy problem; large $| S_{\max} |$ implies a
difficult problem.  For bounding the running time of range search, we require
one more notion of difficulty, related to how $| S_{\max} |$ changes due to
changes in the range $[l, u]$.

@@

When the $\alpha$-expansion of the set $S_{\max}$ is approximately the same size
as $S_{\max}$, then the problem would not be significantly more difficult if the
range $[l, u]$ was increased slightly.  Using these notions, then, we may now
bound the running time of range search.

@@

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime}, we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
proofs, then, our only task is to bound the maximum size of the reference set,
$|R^*|$.

@@

This set is necessarily a subset of $S^{\alpha}[p_q]$; by
assumption, the number of points in this set is bounded above by $|S_{\max}| +
C$.  We may then conclude that $|R^*| \le |S_{\max}| + C$.  By taking the
maximum of the sizes of $|R^*|$ in both cases above, we obtain the statement of
the theorem.
\end{proof}

### 2015-01-08 bounds.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
%Numerous machine learning algorithms contain pairwise statistical problems at
%their core. Often, tree structures are used to solve these problems efficiently.
%Dual-tree algorithms using cover trees have rigorous worst-case runtime
%guarantees for some of these problems. In this paper, we present a {\em
%problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
%the cover tree, separating out the problem-dependent and the problem-independent
%elements. This allows us to just plug in bounds for the problem-dependent
%elements to get runtime guarantees for dual-tree algorithms for any pairwise
%statistical problem. We demonstrate this plug-and-play procedure for
%nearest-neighbor search and approximate kernel density estimation to get
%improved runtime guarantees.  Under some assumptions, we also present the first
%linear runtime guarantee for dual-tree based range search.
Numerous machine learning algorithms contain pairwise statistical problems at
their core -- i.e. tasks that require computations over all pairs of input points if implemented directly. 
Often, tree structures are used to solve these problems efficiently. Dual-tree algorithms 
can efficiently solve or approximate many of these problems.
Using cover trees, rigorous worst-case runtime
guarantees have been proven for some of these algorithms. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
the cover tree, separating out the problem-dependent and the problem-independent
elements. This allows us to plug in bounds for the problem-dependent
elements to get runtime guarantees for dual-tree algorithms for any pairwise
statistical problem without re-deriving the entire proof. We demonstrate this plug-and-play procedure for
nearest-neighbor search and approximate kernel density estimation to get
improved runtime guarantees.  Under mild assumptions, we also present the first
linear runtime guarantee for dual-tree based range search.
\end{abstract}

@@

\citet{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Instead of
building a tree on the reference set and searching with each query point
separately, Gray and Moore suggest also building a query tree and traversing
both the query and reference trees simultaneously (a {\it dual-tree traversal},
from which the class of algorithms takes its name).

@@

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \citep{gray2003nonparametric}, mean shift \citep{wang2007fast},
minimum spanning tree calculation \citep{march2010euclidean}, $n$-point
correlation function estimation \citep{march2012fast}, max-kernel search
\citep{curtin2013fast}, particle smoothing \citep{klaas2006fast}, variational
inference \citep{amizadeh2012variational}, and range search \citep{nbody}, to name
a few.

Some of these algorithms are derived using the cover tree \citep{langford2006}, a
data structure with beneficial theoretical qualities.  Dual-tree all
nearest-neighbor search and approximate kernel density estimation with cover
trees have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009}; minimum
spanning tree calculation scales as $O(N \log N)$ \citep{march2010euclidean}.
Other problems have similar worst-case guarantees \citep{curtin2014dual,
march2013multi}.

>>>>

\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
%Numerous machine learning algorithms contain pairwise statistical problems at
%their core. Often, tree structures are used to solve these problems efficiently.
%Dual-tree algorithms using cover trees have rigorous worst-case runtime
%guarantees for some of these problems. In this paper, we present a {\em
%problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
%the cover tree, separating out the problem-dependent and the problem-independent
%elements. This allows us to just plug in bounds for the problem-dependent
%elements to get runtime guarantees for dual-tree algorithms for any pairwise
%statistical problem. We demonstrate this plug-and-play procedure for
%nearest-neighbor search and approximate kernel density estimation to get
%improved runtime guarantees.  Under some assumptions, we also present the first
%linear runtime guarantee for dual-tree based range search.
Numerous machine learning algorithms contain pairwise statistical problems at
their core -- i.e. tasks that require computations over all pairs of input points if implemented directly. 
Often, tree structures are used to solve these problems efficiently. Dual-tree algorithms 
can efficiently solve or approximate many of these problems.
Using cover trees, rigorous worst-case runtime
guarantees have been proven for some of these algorithms. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
the cover tree, separating out the problem-dependent and the problem-independent
elements.  This allows us to just plug in bounds for the problem-dependent
elements to get runtime guarantees for dual-tree algorithms for any pairwise
statistical problem without re-deriving the entire proof. We demonstrate this plug-and-play procedure for
nearest-neighbor search and approximate kernel density estimation to get
improved runtime guarantees.  Under mild assumptions, we also present the first
linear runtime guarantee for dual-tree based range search.
\end{abstract}

@@

Building upon this intuition, \citet{nbody} generalized the fast multipole
method from computational physics to obtain dual-tree algorithms.  These are
extremely useful when there are large query sets, not just a few query points.
Instead of building a tree on the reference set and searching with each query
point separately, Gray and Moore suggest also building a query tree and
traversing both the query and reference trees simultaneously (a {\it dual-tree
traversal}, from which the class of algorithms takes its name).

@@

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \citep{gray2003nonparametric}, mean shift \citep{wang2007fast},
minimum spanning tree calculation \citep{march2010euclidean}, $n$-point
correlation function estimation \citep{march2012fast}, max-kernel search
\citep{curtin2013fast}, particle smoothing \citep{klaas2006fast}, variational
inference \citep{amizadeh2012variational}, range search \citep{nbody}, and
embedding techniques \cite{maaten2014accelerating}, to name a few.

Some of these algorithms are derived using the cover tree \citep{langford2006}, a
data structure with compelling theoretical qualities.  When cover trees are
used, Dual-tree all-nearest-neighbor search and approximate kernel density
estimation have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009};
minimum spanning tree calculation scales as $O(N \log N)$
\citep{march2010euclidean}.  Other problems have similar worst-case guarantees
\citep{curtin2014dual, march2013multi}.
